A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-07 podc_paper/algorithm-2.tex
minor changes
working on GC; just realized there is a problem that splits might not get installed even after p tries at nodes other than the root

@@

	\gc{
\Function{boolean}{GCRefresh}{\typ{Node} \var{v}}
    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}, with GC}
	\State $\var{T} \assign\ \var{v}.\fld{blocks}$\label{readTGC}
    \State $\var{h} \assign\ \Call{MaxBlock}{T}.\fld{index}$ \label{readHeadGC}
	\If{\var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
		\State \linecomment{garbage collection phase}
		\State \var{max} \assign 0
		\For {$k \assign 1..p$}
			\State $\var{max} \assign \max(\var{max}, \var{last}[k])$
		\EndFor
		\State \Call{Help}{}
		\State \Return $\Call{Discard}{\var{root}, \var{max}-1}$\label{discard}
    \Else\ \linecomment{do a standard refresh}
        \State $T' \assign \Call{RefreshUpdate}{T, v}$\label{RU1}
        \If{$T' = \nl$} \Return{\tr}
    	\Else\ \Return $\Call{CAS}{\var{v}.\fld{blocks}, T, T'}$\label{casGC}
		\EndIf
	\EndIf
%    \State \Call{Advance}{\var{v, h}}\label{advanceGC}
\EndFunction{GCRefresh}
}

@@

\gc{
\Function{void}{SetSuper}{\typ{Node} \var{v}, \typ{RBT} $T$} 
	\State \linecomment{set \fld{super} field of last block in $T$}
	\State \linecomment{and in $v$'s children's RBTs}
	\State $\var{lastBlock} \assign \Call{MaxBlock}{T}$
	\If{$v\neq \var{root}$}	
		\State \var{sup} \assign\ \Call{MaxBlock}{\var{v}.\fld{parent.blocks}}.\fld{index}
		\State \Call{CAS}{\var{lastBlock}.\fld{super}, \nl, \var{sup}}
	\EndIf
	\ForEach{\fld{dir} in \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
    	\State \var{lastChildBlock} \assign\ \Call{MaxBlock}{\var{v}.\fld{dir}.\fld{blocks}}
		\State \Call{CAS}{\var{lastChildBlock}.\fld{super}, \nl, \var{lastBlock}.\fld{index}}
    \EndFor \label{endHelpChild1GC}
\EndFunction{SetChildrenSuper}
}

@@

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
    \gc{
    \State \var{new}.\eleft \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC}
    \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
    \State \var{new}.\fld{index} \assign \var{i}
    }
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}
\end{algorithmic}
\end{minipage}
\caption{Bounded-space queue implementation.\label{pseudocode1GC}}
\end{figure}

@@

\begin{invariant}
For each RBT $T$, 
$T$ is non-empty and the blocks of $T$ have consecutive \fld{index} fields.
\end{invariant}
\begin{proof}
The claim is true initially, since each RBT initially contains only the empty block with \fld{index} 0.
We show that each time a new RBT is constructed, it preserves the invariant.
When a new RBT is created by inserting an element into an existing RBT, \Cref{RBTinsertions}
implies that the invariant is preserved.

@@

Thus, calls to \op{MaxBlock} are well-defined.  Also, this means that the sequence of blocks added to $v.\fld{blocks}$ have consecutive \fld{index} values, starting from 1, so the references to $v.\fld{blocks}[i]$ (meaning the block in $v$'s RBT with \fld{index} $i$) are well-defined.

\begin{corollary}\label{maxNonDecreasing}
The maximum \fld{index} in $v$'s RBT is non-decreasing over time.
\end{corollary}
\begin{proof}
The CAS steps (on line \ref{casGC} and \ref{splitCASGC}) that update $v.\fld{blocks}$ always replace 
a tree $T$ by a tree that is built from $T$ by inserting an element or removing the elements with 
the smallest \fld{index} values.
\end{proof}

@@

\Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} hold true for the modified algorithm.  Their proofs are identical to those given in Section \ref{sec::basicProperties}.

>>>>

\here{Walk reader through the code (and how it is different from basic versino)}

\op{MaxBlock} is a query operation on a RBT that returns the block with the maximum \fld{index}.
We could keep a pointer to the maximum block in the root of the RBT so that \op{MaxBlock}
can be done in constant time.

@@

	\gc{
\Function{boolean}{GCRefresh}{\typ{Node} \var{v}}
    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}, with GC}
	\State $\var{T} \assign\ \var{v}.\fld{blocks}$\label{readTGC}
    \State $\var{h} \assign\ \Call{MaxBlock}{T}.\fld{index}$ \label{readHeadGC}
	\If{\var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
		\State \linecomment{garbage collection phase}
		\State \var{max} \assign 0
		\For {$k \assign 1..p$}
			\State $\var{max} \assign \max(\var{max}, \var{last}[k])$
		\EndFor
		\State \Call{Help}{}
		\State \Return $\Call{Discard}{\var{root}, \var{max}-1}$\label{discard}
    \Else\ \linecomment{do a standard refresh}
        \State $T' \assign \Call{RefreshUpdate}{T, v}$\label{RU1}
        \If{$T' = \nl$} \Return{\tr}\label{emptyblockGC}
    	\Else\ \Return $\Call{CAS}{\var{v}.\fld{blocks}, T, T'}$\label{casGC}
		\EndIf
	\EndIf
%    \State \Call{Advance}{\var{v, h}}\label{advanceGC}
\EndFunction{GCRefresh}
}

@@

\gc{
\Function{void}{SetSuper}{\typ{Node} \var{v}, \typ{RBT} $T$} 
	\State \linecomment{set \fld{super} field of last block in $T$}
	\State \linecomment{and in $v$'s children's RBTs}
	\State $\var{lastBlock} \assign \Call{MaxBlock}{T}$
	\If{$v\neq \var{root}$}	
		\State \var{sup} \assign\ \Call{MaxBlock}{\var{v}.\fld{parent.blocks}}.\fld{index}
		\State \Call{CAS}{\var{lastBlock}.\fld{super}, \nl, \var{sup}}
	\EndIf
	\ForEach{\fld{dir} in \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
    	\State \var{lastChildBlock} \assign\ \Call{MaxBlock}{\var{v}.\fld{dir}.\fld{blocks}}
		\State \Call{CAS}{\var{lastChildBlock}.\fld{super}, \nl, \var{lastBlock}.\fld{index}}
    \EndFor \label{endHelpChild1GC}
\EndFunction{SetSuper}
}

@@

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
    \gc{
    \State \var{new}.\eleft \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC}
    \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
    \State \var{new}.\fld{index} \assign \var{i}
    }
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}\label{CBnullGC}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}
\end{algorithmic}
\end{minipage}
\caption{Bounded-space queue implementation.\label{pseudocode1GC}}
\end{figure}

@@

\begin{invariant}\label{nonemptyConsec}
For each RBT $T$, 
$T$ is non-empty and the blocks of $T$ have consecutive \fld{index} fields.
\end{invariant}
\begin{proof}
The claim is true initially, since each RBT initially contains only the empty block with \fld{index} 0.
We show that each time a new RBT is constructed, it preserves the invariant.
When a new RBT is created by inserting an element into an existing RBT, \Cref{RBTinsertions}
implies that the invariant is preserved.

@@

Since RBTs are always non-empty, calls to \op{MaxBlock} are well-defined.  
Throughout the proof, we use $v.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
in $v$'s tree at some time during the execution.
It follows from \Cref{RBTinsertions} and \Cref{nonemptyConsec} that this block is unique, if 
it exists.
We also use this notation in the code to indicate that a search of the RBT should be performed for the block
with \fld{index} $b$.
We describe later what should happen if the block is not found by the search because it has been removed
by a GC phase.
\here{remember to come back to this point}

\begin{corollary}\label{maxNonDecreasing}
The maximum \fld{index} in $v$'s RBT is non-decreasing over time.
\end{corollary}
\begin{proof}
The CAS steps (on line \ref{casGC} and \ref{splitCASGC}) that update $v.\fld{blocks}$ always replace 
a tree $T$ by a tree that is built from $T$ by inserting an element and/or removing the elements with 
the smallest \fld{index} values, without emptying the tree.  Thus, the maximum \fld{index} in the tree cannot decrease.
\end{proof}

@@

\Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} 
hold true for the modified algorithm.  Their proofs are identical to those given in 
Section \ref{sec::basicProperties} since they only depend on \Cref{lem::headProgress} 
and the definition of subblocks given in (\ref{defsubblock}).
The following pair of lemmas are analogous to \Cref{successfulRefresh} for the modified \op{Refresh} routine.
We say that an operation has been \emph{propagated} to node $v$ in configuration $C$ if
some block containing the operation has been in $v.blocks$ at or before $C$.

\begin{lemma}\label{nullRefreshGC}
Suppose $v.\fld{blocks}=T$ at some configuration $C$ and \op{RefreshUpdate}($T,v$) is invoked after $C$ and
returns \nl.  Then, all operations that were propagated to $v$'s
children before $C$ have been propagated to $v$ before $C$.
\end{lemma}
\begin{proof}
Suppose an operation $op$ has been propagated to $v$'s child (without loss of generality, $v.\fld{left}$) before $C$.
Then, for some $i$, a block $B$ with \fld{index} $i$ has been in $v.\fld{left.blocks}$ before $C$.
By Lemma \ref{maxNonDecreasing}, 
the maximum \fld{index} in $v.\fld{left.blocks}$ is at least $i$ at all times during the \op{RefreshUpdate}.
So, the value of $\var{new}.\eleft$ chosen at line \ref{createEndLeftGC} of the \op{RefreshUpdate}'s call
to \op{CreateBlock} is at least $i$.

Let $h$ be the value set at line \ref{refreshReadMax} of the \op{RefreshUpdate}.
The \op{RefreshUpdate} returns \nl\ at line \ref{addOPGC}
because the \op{CreateBlock}$(v,h)$  at line \ref{CBnullGC} returns \nl.
By reasoning identical to the last paragraph of \Cref{successfulRefresh},
it follows from the fact that $\var{num\sub{enq}}+\var{num\sub{deq}}=0$ that
the blocks $v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$
contain no operations.
Since each block contains at least one operation by \Cref{blockNotEmpty}
these ranges must be empty, so $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft\geq i$.
This implies that the block $B$ containing $op$ is a subblock of some block that has appeared in $v.\fld{blocks}$, so $op$ has been propagated to $v$.
\end{proof}

\begin{lemma}\label{successfulRefreshGC}
If a \op{CAS} at line \ref{casGC} or \ref{splitCASGC} modifies $v.\fld{blocks}$ from $T$ to $T'$,
then all operations that were propagated to $v$'s children before $v.\fld{blocks}$ was set to $T$
have been propagated to $v$ in the configuration after the CAS.
\end{lemma}
\begin{proof}
Suppose an operation $op$ has been propagated to $v$'s child (without loss of generality, $v.\fld{left}$) before $v.\fld{blocks}$ was set to $T$.
Then, for some $i$, a block $B$ with \fld{index} $i$ that contains $op$
has been in $v.\fld{left.blocks}$ before $v.\fld{blocks}$ was set to $T$.
Let $c$ be the \op{CAS} instruction that changes $v.\fld{blocks}$ from $T$ to $T'$.
If line \ref{casGC} performs $c$, then the \op{RefreshUpdate} $R$ at line \ref{RU1} that creates
$T'$ began after $T$ was read from $v.\fld{blocks}$ at line \ref{readTGC}.
Otherwise, line \ref{splitCASGC} performs $c$ and the \op{RefreshUpdate} $R$ at line \ref{RU2},
which is invoked after $T$ was read from $v.\fld{blocks}$ at line \ref{readT2GC},
either returned \nl\ or $T'$.

If the \op{RefreshUpdate} $R$ returned \nl, the lemma follows from \Cref{nullRefreshGC}.
So, assume \op{RefreshUpdate} $R$ returned $T'$.
Then $T'$ contains a newly inserted block created by the call
to \op{CreateBlock} on line \ref{invokeCreateBlockGC}.
Since $R$ was invoked after $T$ was stored in $v.\fld{blocks}$, 
the maximum \fld{index} in $v.\fld{left.blocks}$ is at least $i$ throughout $R$, by \Cref{maxNonDecreasing}.
Thus. the \eleft\ field of the new block added to $T'$ by $R$'s call to \op{CreateBlock} is
set to a value greater than or equal to $i$ at line \ref{createEndLeftGC}.
By the definition of subblocks, it follows that the block $B$ containing $op$ is a subblock of a block
that has been in $v.blocks$, as required.
\end{proof}

This allows us to show that a double \op{Refresh} propagates operations up the tree, 
as in \Cref{lem::doubleRefresh}.

\begin{lemma}\label{lem::doubleRefreshGC}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{GCRefresh}($v$) by the same process.
All operations propagated to $v$'s children when $R_1$ begins
are propagated to $v$ when $R_2$ terminates.
\end{lemma}
\begin{proof}
\end{proof}

### 2023-01-07 podc_paper/correctness.tex
working on GC; just realized there is a problem that splits might not get installed even after p tries at nodes other than the root

\begin{lemma}\label{successfulRefresh}
Let $R$ be a call to \op{Refresh}($v$) that performs a successful \op{CAS} on line \ref{cas} (or terminates at line \ref{addOP}).
In the configuration after that CAS (or termination, respectively), $v$ contains all operations that $v$'s children contained 
when $R$ executed line~\ref{readHead}.
\end{lemma}
\begin{proof}
Suppose $v$'s child (without loss of generality, $v.\fld{left}$) contained an operation $op$ 
in the configuration $C$ immediately before $R$ executed line \ref{readHead}.
Then some block $B$ in $v.\fld{left.blocks}[i]$ contains $op$.
By \Cref{nonDecreasingHead} and \Cref{lem::headProgress}, the value of $childHead$ that $R$ reads from
$v.\fld{left.head}$ in line \ref{readChildHead} is at least $i$.
If it is equal to $i$, $R$ calls \op{Advance} at line \ref{helpAdvance}, which ensures that 
$v.\fld{left.head} > i$.
Then, $R$ calls \op{CreateBlock}($v,h$) in line \ref{invokeCreateBlock}, where $h$ is the value $R$ reads at line \ref{readHead}.
\op{CreateBlock} reads a value greater than $i$ from $v.\fld{left.head}$ at line \ref{createEndLeft}.
Thus, $new.\eleft \geq i$.  We consider two cases.

@@

\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $v$'s children to promote, $op$ is already in $v$.
We formalize this intuition.
The value computed at line \ref{createSumEnq} is
\begin{eqnarray*}
\var{num\sub{enq}} 
&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} - v.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
&&\mbox{ }- v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}
\end{eqnarray*}
It follows from \Cref{lem::sum} that $num\sub{enq}$ is the total number of enqueues contained in the blocks
$v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$.
Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By \Cref{blockNotEmpty}, this means the ranges of blocks are empty, so that $v.\fld{blocks}[h-1].\eleft \geq new.\eleft \geq i$.
Hence, $B$ is already a subblock of some block in $v$, so $v$ contains $op$.
\end{proof}

@@

\begin{lemma}\label{lem::doubleRefresh}
Consider two consecutive terminating calls $R_1$, $R_2$ to \nf{Refresh}($v$) by the same process.
When $R_2$ terminates, $v$ contains all operations that $v$'s children contained when $R_1$ was invoked.
\end{lemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
from \Cref{successfulRefresh}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $v.\head$ at line \ref{readHead}.
By \Cref{lem::headInc}, $h_2>h_1$.
By \Cref{lem::headProgress}, $v.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
Since $R_2$ fails its \op{CAS} on $v.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
a successful \op{CAS} on $v.blocks[h_2]$ before $R_2$'s \op{CAS}.
$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $v.\head$ and the value of $v.\head$ is non-decreasing, by \Cref{nonDecreasingHead}.
Thus, all operations contained in $v$'s children when $R_1$ begins
are also contained in $v$'s children when $R_3$ later executes line \ref{readHead}.
By \Cref{successfulRefresh}, these operations are contained in $v$ when $R_3$ performs its successful \op{CAS},
which is before $R_2$ terminates.
\end{proof}

>>>>

\begin{lemma}\label{successfulRefresh}
Let $R$ be a call to \op{Refresh}($v$) that performs a successful \op{CAS} on line \ref{cas} (or terminates at line \ref{addOP}).
In the configuration after that CAS (or termination, respectively), $v$ contains all operations that $v$'s children contained 
when $R$ executed line~\ref{readHead}.
\end{lemma}
\begin{proof}
Suppose $v$'s child (without loss of generality, $v.\fld{left}$) contained an operation $op$ 
in the configuration $C$ immediately before $R$ executed line \ref{readHead}.
Then for some $i$, the block $B$ in $v.\fld{left.blocks}[i]$ contains $op$.
By \Cref{nonDecreasingHead} and \Cref{lem::headProgress}, the value of $childHead$ that $R$ reads from
$v.\fld{left.head}$ in line \ref{readChildHead} is at least $i$.
If it is equal to $i$, $R$ calls \op{Advance} at line \ref{helpAdvance}, which ensures that 
$v.\fld{left.head} > i$.
Then, $R$ calls \op{CreateBlock}($v,h$) in line \ref{invokeCreateBlock}, where $h$ is the value $R$ reads at line \ref{readHead}.
\op{CreateBlock} reads a value greater than $i$ from $v.\fld{left.head}$ at line \ref{createEndLeft}.
Thus, $new.\eleft \geq i$.  We consider two cases.

@@

\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $v$'s children to promote, $op$ is already in $v$.
We formalize this intuition.
The value computed at line \ref{createSumEnq} is
\begin{eqnarray*}
\var{num\sub{enq}} 
&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} - v.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
&&\mbox{ }- v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}
\end{eqnarray*}
It follows from \Cref{lem::sum} that $num\sub{enq}$ is the total number of enqueues contained in the blocks
$v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$.
Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By \Cref{blockNotEmpty}, this means the ranges of blocks are empty, so that $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft \geq i$.
Hence, $B$ is already a subblock of some block in $v$, so $v$ contains $op$.
\end{proof}

@@

\begin{lemma}\label{lem::doubleRefresh}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($v$) by the same process.
When $R_2$ terminates, $v$ contains all operations that $v$'s children contained when $R_1$ was invoked.
\end{lemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
from \Cref{successfulRefresh}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $v.\head$ at line \ref{readHead}.
By \Cref{lem::headInc}, $h_2>h_1$.
By \Cref{lem::headProgress}, $v.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
Since $R_2$ fails its \op{CAS} on $v.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
a successful \op{CAS} on $v.blocks[h_2]$ before $R_2$'s \op{CAS}.
$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $v.\head$ and the value of $v.\head$ is non-decreasing, by \Cref{nonDecreasingHead}.
Thus, all operations contained in $v$'s children when $R_1$ begins
are also contained in $v$'s children when $R_3$ later executes line \ref{readHead}.
By \Cref{successfulRefresh}, these operations are contained in $v$ when $R_3$ performs its successful \op{CAS},
which is before $R_2$ terminates.
\end{proof}
